Perturbing SU($n$)-symmetric interactions

### 2019-11-09 collective_sun/sun_perturbation.tex
overhauled some notation that will allow finishing the calculation of the second-order effect of two-body perturabtions

\newcommand{\B}{\mathcal{B}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\Z}{\mathcal{Z}}

@@

We wish to determine the effective dynamics induced on the
ground-state manifold $\M_0$ of $H_0$ by weak single- and two-body
perturbations of the form
\begin{align}
  \V_1 \equiv \sum_p v_p V_p = \v v\c\v V,
  &&
  \V_2 \equiv \f12 \sum_{p,q} w_{pq} X_p Y_q
  = \f12 \v X \c\m w \c\v Y,
  \label{eq:perturbations}
\end{align}
where $V_p,X_p,Y_p$ are trace-zero operators on spin $p$, with
e.g.~$V_p$ related to $V_q$ by a permutation of spins $p$ and $q$;
$\v V,\v X,\v Y$ are respectively vectors of all $V_p,X_p,Y_p$; $\v v$
and $\m w$ are respectively a vector and matrix of all $v_p,w_{pq}$;
we enforce $w_{pq}=w_{qp}$ and $w_{pp}=0$; and we use a dot ($\c$) to
denote an inner product between vectors and matrices as appropriate.

@@

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we use the permutational symmetry of the
ground-state manifold to simplify
\begin{align}
  H_1^{(1)} = \sum_p v_p \P_0 V_p \P_0
  = \sum_p v_p \P_0 V_0 \P_0 \FSEQ \bar v V,
  \label{eq:H_1_1}
\end{align}
where $\FSEQ$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, and
\begin{align}
  \bar v \equiv \f1N \sum_p v_p,
  &&
  V \equiv \sum_p V_p.
\end{align}
In order to compute the second order effective Hamiltonian
$H_1^{(2)}$, we first choose an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \p{\v v\c\m h - \v v\oc\v h}\c \v V \ket\psi,
  \label{eq:H_V1_psi_start}
\end{align}
where $\m h$ is a matrix of all coefficients $h_{pq}$ of the
interaction Hamiltonian $H_0$ in \eqref{eq:ints}; $\v h$ is a vector
of the net interaction energies $h_p\equiv\sum_qh_{pq}$ associated
with individual spins $p$; and $\v v\oc\v h$ is the element-wise
(Hadamard) product of $\v v$ and $\v h$, i.e.~a vector with components
$v_kh_k$.  We thus find that the vector $\V_1\ket\psi$ is an
eigenvector of the interaction Hamiltonian $H_0$ if the coefficients
$\v v$ satisfy
\begin{align}
  \m h \c \v v - \v h\oc\v v
  = \p{\m h - \diag\v h}\c\v v
  = \Delta\p{\v v} \v v,
  \label{eq:cond_1}
\end{align}
where $\diag\v h$ is a matrix with $\v h$ on the diagonal and zeroes
everywhere else; and $\Delta\p{\v v}$ is some constant that may depend
on $\v v$.  If the interaction Hamiltonian $H_0$ is translationally
invariant, then all $h_p=h\equiv\EE_q\sp{h_q}$ and
$\v h\oc\v v=h\v v$, so the condition in \eqref{eq:cond_1} simplifies
to
\begin{align}
  \m h \c\v v = \sp{\Delta\p{\v v}+h} \v v.
\end{align}
If we {\it construct} a perturbation $\V_1\p{\v v_\Delta}$ with
coefficients $\v v_\Delta$ that satisfy \eqref{eq:cond_1} with
eigenvalue $\Delta$, then
\begin{align}
  H_0 \V_1\p{\v v_\Delta} \ket\psi
  = \p{E_0 + \Delta} \V_1\p{\v v_\Delta} \ket\psi.
  \label{eq:H_V1_psi}
\end{align}
Interestingly, the energy of the (unnormalized) state
$\V_1\p{\v v_\Delta}\ket\psi$ depends only on the coefficients
$\v v_\Delta$, and is entirely independent of the state $\ket\psi$ or
choice of single-spin trace-zero operator $V_p$ used to build $\V_1$
in \eqref{eq:perturbations}.  Finding operators $\V_1\p{\v v_\Delta}$
that generate eigenvectors of the interaction Hamiltonian $H_0$ when
they act on fully symmetric states $\ket\psi$ thus reduces to finding
eigenvectors $\v v_\Delta$ of the interaction matrix $\m h-\diag\v h$.

@@

We now return to the task of computing the second-order effective
Hamiltonian $H_1^{(2)}$.  Any coefficient vector $\v v$ can be
expanded into its projections $\v v_\Delta$ onto the eigenspace of
vectors satisfying \eqref{eq:cond_1} with eigenvalue $\Delta$,
i.e.~$\v v = \sum_\Delta \v v_\Delta$, which also allows us to expand
\begin{align}
  \V_1\p{\v v} = \sum_\Delta \V_1\p{\v v_\Delta},
\end{align}
where each operator $\V_1\p{\v v_\Delta}$ generates a state with
definite interaction energy $\Delta$ above that of the fully symmetric
manifold $\M_0$.  We can therefore simplify
\begin{align}
  H_1^{(2)}
  = - \P_0 \V_1\p{\v v} \E \V_1\p{\v v} \P_0
  = - \sum_{\Delta>0} \f1{\Delta}
  \P_0 \V_1\p{\v v} \P_\Delta \V_1\p{\v v} \P_0
  = - \sum_{\Delta>0} \f1{\Delta} \P_0 \V_1\p{\v v_\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \V_1\p{\v v_\Delta}^2 \P_0$ is worked out in
Appendix \ref{sec:PVVP}.  In total, we find
\begin{align}
  H_1^{(2)} \FSEQ \f{\tilde v}{N-1} \p{V^2 - N W},
  &&
  \tilde v \equiv \sum_{\Delta>0} \f{\var\p{\v v_\Delta}}{\Delta},
  \label{eq:H_1_2}
\end{align}
where $\FSEQ$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, and
\begin{align}
  \var\p{\v x} \equiv \EE_p\sp{\p{x_p-\EE_q\sp{x_q}}^2}
  = \EE_p\sp{x_p^2} - \EE_q\sp{x_q}^2,
  &&
  W \equiv \sum_p V_p^2,
\end{align}
with $\EE_p\sp{x_p}\equiv\sum_px_p/N$ the mean value in $\v x$.

@@

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} is simply the restriction of $\V_2$ onto
$\M_0$ (see Appendix \ref{sec:PXYP}):
\begin{align}
  H_2^{(1)}
  = \P_0 \V_2 \P_0
  \FSEQ \f12 \bar w \p{\tilde Z - Z},
\end{align}
where in terms of the anti-commutator $\sp{A,B}_+\equiv AB+BA$ and
collective operators $X\equiv\sum_p X_p$ and $Y\equiv\sum_p Y_p$, we
define
\begin{align}
  \bar w \equiv {N\choose 2}^{-1} \sum_{p<q} w_{pq},
  &&
  \tilde Z \equiv \f12 \sp{X,Y}_+,
  &&
  Z \equiv \sum_p \f12 \sp{X_p, Y_p}_+.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V2_psi})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \v X \c \p{\bar{\m h}\oc\m w + \m h\c\m w} \c \v Y \ket\psi
  - \v\1 \c \p{\m h\oc\m w} \c \v Z \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where $\v\1$ is an $N$-component vector of ones;
\begin{align}
  \bar{\m h}
  \equiv \m h - \f12\p{\v h\otimes\v\1 + \v\1\otimes\v h}
\end{align}
is a modified interaction matrix with entries $h_{pq}-\p{h_p+h_q}/2$;
and $\m x\oc\m y$ is an element-wise (Hadamard) product of $\m x$ and
$\m y$, with entries $x_{pq}y_{pq}$.  We thus find that the vector
$\V_2\ket\psi$ is an eigenvector of the interaction Hamiltonian $H_0$
if the matrix $\m w$ of all $w_{pq}$ satisfies
\begin{align}
  \bar{\m h}\oc\m w + \m h\c\m w = \f12 \Delta\p{\m w} \m w,
  \label{eq:cond_2}
\end{align}
for some constant $\Delta\p{\m w}$ that may depend on $\m w$.
Remembering that the matrices $\m h$ and $\m w$ are both symmetric
(under transposition) with zeros on the diagonal, the diagonal
components of the condition in \eqref{eq:cond_2} automatically enforce
that the last term ($\sim\v Z$) in \eqref{eq:H_V2_psi} vanishes.

We can recast the condition in \eqref{eq:cond_2} as an eigenvalue
problem by writing the coefficient matrix $\m w$ as a vector
$\v{\m w}$ in $\RR^N\times\RR^N$,
\begin{align}
  \m w = \sum_{p,q} w_{pq} \op{p}{q}
  \to \v{\m w} \equiv \sum_{p,q} w_{pq} \ket{pq},
\end{align}
and defining the matrix
\begin{align}
  \check{\m h}
  \equiv \sum_{p,q} \bar h_{pq} \op{pq}
  + \sum_{p,q,\ell} h_{pq} \op{p\ell}{q\ell},
  \label{eq:h_super_mat}
\end{align}
in terms of which the condition in \eqref{eq:cond_2} becomes the
eigenvalue equation
\begin{align}
  \check{\m h} \c \v{\m w} = \f12 \Delta\p{\m w} \v{\m w}.
  \label{eq:cond_2_eig}
\end{align}
The matrix $\check{\m h}$ in \eqref{eq:h_super_mat} can be written in
the block-diagonal form
\begin{align}
  \check{\m h} = \sum_\ell \m h_\ell \otimes \op{\ell},
  &&
  \m h_\ell
  \equiv \sum_{p,q}\p{\delta_{pq} \bar h_{p\ell} + h_{pq}} \op{p}{q},
\end{align}
which reduces the eigenvalue equation in \eqref{eq:cond_2_eig} to the
$N$ smaller, independent eigenvalue equations
\begin{align}
  \m h_\ell \c \v{\m w}_\ell = \f12 \Delta\p{\m w} \v w_\ell,
  &&
  \v w_\ell \equiv \sum_p w_{p\ell} \ket{p}.
  \label{eq:cond_2_eig_block}
\end{align}
We can (numerically) diagonalize each of the blocks $\m h_\ell$, and
classify their eigenvectors $\v w_\ell$ by the corresponding
eigenvalues $\Delta\p{\m w}/2$.  ``Total'' eigenvectors $\v{\m w}$ of
$\check{\m h}$ are then built from eigenvectors $\v w_\ell$ within
each block $\m h_\ell$ that share the same eigenvalue
$\Delta\p{\m w}/2$.  The task of constructing eigenvectors
$\V_2\p{\m w_\Delta}\ket\psi$ of $H_0$ and determining their energy
thus reduces to finding the eigenvectors and eigenvalues of $N$
matrices $\m h_\ell$ of size $N\times N$.

@@

We wish to simplify
\begin{align}
  H_0 \V_1 \ket\psi
  = \sum_{\substack{k\\p<q}} h_{pq} v_k \Pi_{pq} V_k \ket\psi,
  \label{eq:H_V1_psi_start_apndx}
\end{align}
which has terms with $k\in\set{p,q}$, and terms with
$k\notin\set{p,q}$.  In the case of $k\notin\set{p,q}$, the
permutation operator $\Pi_{pq}$ commutes with $V_k$ and annihilates on
$\ket\psi$, leaving us the sum
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}} h_{pq}
  = \f12 \sum_{\substack{p,q\\k\notin\set{p,q}}} h_{pq}
  = \f12 \sum_{p,q} h_{pq}
  - \f12 \sum_{\substack{p,q\\k\in\set{p,q}}} h_{pq}
  = E_0 - h_k,
\end{align}
where we have used the facts that $h_{pq}=h_{qp}$ and $h_{pp}=0$.  The
terms in \eqref{eq:H_V1_psi_start_apndx} with $k\notin\set{p,q}$ are
then
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}}
  h_{pq} v_k \Pi_{pq} V_k \ket\psi
  = \sum_k \p{E_0 - h_k} v_k V_k \ket\psi
  = E_0 \V_1 \ket\psi - \v h\oc\v v \c\v V \ket\psi,
\end{align}
where $\v h\oc\v v$ is an element-wise (Hadamard) product of $\v h$
and $\v v$.  The terms with $k\in\set{p,q}$, meanwhile, are
\begin{align}
  \sum_{\substack{p<q\\k\in\set{p,q}}}
  h_{pq} v_k \Pi_{pq} V_k \ket\psi
  = \sum_{k<q} h_{kq} v_k V_q \ket\psi
  + \sum_{p<k} h_{pk} v_k V_p \ket\psi
  = \sum_{k,q} h_{kq} v_k V_q \ket\psi
  = \v v\c\m h\c\v V \ket\psi,
\end{align}
which implies
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \p{\v v\c\m h - \v v\oc\v h}\c \v V \ket\psi.
\end{align}

@@

We wish to simplify
\begin{align}
  H_0 \V_2 \ket\psi
  = \f12 \sum_{\substack{k,\ell\\p<q}} w_{k\ell} h_{pq}
  \Pi_{pq} X_k Y_\ell \ket\psi,
  \label{eq:H_V2_psi_start}
\end{align}
which has one term with $\set{p,q}=\set{k,\ell}$, terms with
$p,q\notin\set{k,\ell}$, and terms with $p,q\in\set{k,\ell}$.  The
permutation operator $\Pi_{pq}$ acts trivially on
$X_k Y_\ell \ket\psi$ when $\set{p,q}=\set{k,\ell}$ or
$p,q\notin\set{k,\ell}$, leaving a sum over $p,q$ of the form
\begin{align}
  h_{k\ell} + \sum_{\substack{p<q\\p,q\notin\set{k,\ell}}} h_{pq}
  = \sum_{p<q} h_{pq}
  - \sum_{\substack{p\in\set{k,\ell}\\q\notin\set{k,\ell}}} h_{pq}
  = E_0 + 2 h_{k\ell} - h_k - h_\ell.
\end{align}
The corresponding terms in \eqref{eq:H_V2_psi_start} with
$p,q\notin\set{k,\ell}$ and $\set{p,q}=\set{k,\ell}$ are then
\begin{align}
  E_0 \V_2 \ket\psi + \sum_{k,\ell}
  \sp{h_{k\ell} - \f12\p{h_k+h_\ell}} w_{k\ell} X_k Y_\ell \ket\psi
  = E_0 \V_2 \ket\psi + \v X \c \p{\bar{\m h}\oc\m w} \c \v Y \ket\psi,
\end{align}
where $\bar{\m h}\oc\m w$ is an element-wise (Hadamard) product
between $\bar{\m h}$ and $\m w$,
\begin{align}
  \bar{\m h} \equiv \m h - \f12\p{\v h\otimes\v\1 + \v\1\otimes\v h},
\end{align}
and $\v\1$ is a vector of ones.  The terms in
\eqref{eq:H_V2_psi_start} with only one of $p,q\in\set{k,\ell}$
simplify to
\begin{align}
  \f12 \sum_{\substack{k,\ell\\q\notin\set{k,\ell}}} w_{k\ell}
  \p{h_{kq} X_q Y_\ell + h_{\ell q} X_k Y_q} \ket\psi
  &= \sum_{k,\ell,q} w_{k\ell} h_{kq} X_q Y_\ell \ket\psi
  - \sum_{k,\ell} w_{k\ell} h_{k\ell} Z_k \ket\psi \\
  &= \v X \c\m h \c\m w\c\v Y \ket\psi
  - \v\1 \c \p{\m h\oc\m w}\c \v Z \ket\psi,
\end{align}
where
\begin{align}
  Z_k \equiv \f12 \sp{X_k, Y_k}_+,
\end{align}
and $\v Z$ is a vector of all $Z_p$.  In total,
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \v X \c \p{\bar{\m h}\oc\m w + \m h\c\m w} \c \v Y \ket\psi
  - \v\1 \c \p{\m h\oc\m w}\c \v Z \ket\psi.
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Single-body product}
\label{sec:PVVP}

>>>>

\newcommand{\B}{\mathcal{B}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\Z}{\mathcal{Z}}

@@

We wish to determine the effective dynamics induced on the
ground-state manifold $\M_0$ of $H_0$ by weak single- and two-body
perturbations of the form
\begin{align}
  \V_1 \equiv \sum_p v_p X_p = \v v\c\v X,
  &&
  \V_2 \equiv \f12 \sum_{p,q} w_{pq} X_p Y_q
  = \f12 \v X \c\m w \c\v Y,
  \label{eq:perturbations}
\end{align}
where $X_p,Y_p$ are trace-zero operators on spin $p$, with e.g.~$X_p$
related to $X_q$ by a permutation of spins $p$ and $q$; $\v X,\v Y$
are respectively vectors of all $X_p,Y_p$; $\v v$ and $\m w$ are
respectively a vector and matrix of all $v_p,w_{pq}$; we enforce
$w_{pq}=w_{qp}$ and $w_{pp}=0$; and we use a dot ($\c$) to denote an
inner product between vectors and matrices as appropriate.

@@

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we use the permutational symmetry of the
ground-state manifold to simplify
\begin{align}
  H_1^{(1)} = \sum_p v_p \P_0 X_p \P_0
  = \sum_p v_p \P_0 X_0 \P_0 \FSEQ \bar v \X,
  \label{eq:H_1_1}
\end{align}
where $\FSEQ$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, and
\begin{align}
  \bar v \equiv \f1N \sum_p v_p,
  &&
  \X \equiv \sum_p X_p.
\end{align}
In order to compute the second order effective Hamiltonian
$H_1^{(2)}$, we first choose an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1 \ket\psi
  &= E_0 \V_1 \ket\psi
  + \p{\sum_{p,q} h_{pq} v_p X_q - \sum_q v_q h_q X_q} \ket\psi \\
  &= E_0 \V_1 \ket\psi
  + \v v\c \p{\m h - \diag\v h}\c \v X \ket\psi,
  \label{eq:H_V1_psi_start}
\end{align}
where $\m h$ is a matrix of all coefficients $h_{pq}$ of the
interaction Hamiltonian $H_0$ in \eqref{eq:ints}; $\v h$ is a vector
of the net interaction energies $h_q\equiv\sum_ph_{pq}$ associated
with individual spins $p$; and $\diag\v h$ is a matrix with $\v h$ on
the diagonal and zeros everywhere else.  We thus find that the vector
$\V_1\ket\psi$ is an eigenvector of the interaction Hamiltonian $H_0$
if the coefficients $\v v$ satisfy
\begin{align}
  \p{\m h - \diag\v h}\c\v v = \Delta\p{\v v} \v v.
  \label{eq:cond_1}
\end{align}
If we {\it construct} a perturbation $\V_1\p{\v v_\Delta}$ with
coefficients $\v v_\Delta$ that satisfy \eqref{eq:cond_1} with
eigenvalue $\Delta$, then
\begin{align}
  H_0 \V_1\p{\v v_\Delta} \ket\psi
  = \p{E_0 + \Delta} \V_1\p{\v v_\Delta} \ket\psi.
  \label{eq:H_V1_psi}
\end{align}
Interestingly, the energy of the (unnormalized) state
$\V_1\p{\v v_\Delta}\ket\psi$ depends only on the coefficients
$\v v_\Delta$, and is entirely independent of the state $\ket\psi$ or
choice of single-spin trace-zero operator $X_p$ used to build $\V_1$
in \eqref{eq:perturbations}.  Finding operators $\V_1\p{\v v_\Delta}$
that generate eigenvectors of the interaction Hamiltonian $H_0$ when
they act on fully symmetric states $\ket\psi$ thus reduces to finding
eigenvectors $\v v_\Delta$ of the interaction matrix $\m h-\diag\v h$.

@@

We now return to the task of computing the second-order effective
Hamiltonian $H_1^{(2)}$.  Any coefficient vector $\v v$ can be
expanded into its projections $\v v_\Delta$ onto the eigenspace of
vectors satisfying \eqref{eq:cond_1} with eigenvalue $\Delta$,
i.e.~$\v v = \sum_\Delta \v v_\Delta$, which also allows us to expand
\begin{align}
  \V_1\p{\v v} = \sum_\Delta \V_1\p{\v v_\Delta},
\end{align}
where each operator $\V_1\p{\v v_\Delta}$ generates a state with
definite interaction energy $\Delta$ above that of the fully symmetric
manifold $\M_0$.  We can therefore simplify
\begin{align}
  H_1^{(2)}
  = - \P_0 \V_1\p{\v v} \E \V_1\p{\v v} \P_0
  = - \sum_{\Delta>0} \f1{\Delta}
  \P_0 \V_1\p{\v v} \P_\Delta \V_1\p{\v v} \P_0
  = - \sum_{\Delta>0} \f1{\Delta} \P_0 \V_1\p{\v v_\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \V_1\p{\v v_\Delta}^2 \P_0$ is worked out in
Appendix \ref{sec:PXXP}.  In total, we find
\begin{align}
  H_1^{(2)} \FSEQ \f{\tilde v}{N-1} \p{\X^2 - N \X^{(2)}},
  &&
  \tilde v \equiv \sum_{\Delta>0} \f{\var\p{\v v_\Delta}}{\Delta},
  \label{eq:H_1_2}
\end{align}
where $\FSEQ$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, and
\begin{align}
  \X^{(2)} \equiv \sum_p X_p^2,
  &&
  \var\p{\v x} \equiv \EE_p\sp{\p{x_p-\EE_q\sp{x_q}}^2}
  = \EE_p\sp{x_p^2} - \EE_q\sp{x_q}^2,
\end{align}
with $\EE_p\sp{x_p}\equiv\sum_px_p/N$ the mean value in $\v x$.

@@

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} is simply the restriction of $\V_2$ onto
$\M_0$ (see Appendix \ref{sec:PXYP}):
\begin{align}
  H_2^{(1)}
  = \P_0 \V_2 \P_0
  \FSEQ \f12 \bar w \p{\X\odot\Y - \O_{X,Y}},
\end{align}
where $A\odot B\equiv\p{AB+BA}/2$ is a symmetrized product of $A$ and
$B$; and
\begin{align}
  \bar w \equiv {N\choose 2}^{-1} \sum_{p<q} w_{pq},
  &&
  \O_{X,Y} \equiv \sum_p X_p \odot Y_p.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V2_psi})
\begin{align}
  H_0 \V_2 \ket\psi
  &= E_0 \V_2 \ket\psi
  + \sum_{k,\ell} \bar h_{k\ell} w_{k\ell} X_k Y_\ell \ket\psi
  + \sum_{k,\ell,q} h_{kq} w_{q\ell} X_k Y_\ell \ket\psi
  - \sum_{k,\ell} w_{k\ell} h_{k\ell} X_k \odot Y_k \ket\psi \\
  &= E_0 \V_2 \ket\psi
  + \v X \c \p{\bar{\m h}\oc\m w + \m h\c\m w} \c \v Y \ket\psi
  - \v\1 \c \p{\m h\oc\m w} \c \v O_{X,Y} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where $\m w,\m h,\bar{\m h}$ are matrices of all
$w_{k\ell},h_{k\ell},\bar h_{k\ell}$, with
\begin{align}
  \bar h_{k\ell} \equiv h_{k\ell} - \f12\p{h_k + h_\ell},
  &&
  h_k \equiv \sum_\ell h_{k\ell};
\end{align}
$\m x\oc\m y$ is an element-wise (Hadamard) product of $\m x$ and
$\m y$, i.e.~with matrix elements $x_{k\ell} y_{k\ell}$;
$\v\1\equiv\p{1,1,1,\cdots}$ is an $N$-component vector of ones; and
$\v O_{X,Y}$ is a vector with components $\p{X_kY_k+Y_kX_k}/2$, i.e.
\begin{align}
  \v O_{X,Y} \equiv \f12\p{\v X\oc\v Y + \v Y\oc\v X}.
\end{align}
We thus find that the vector $\V_2\ket\psi$ is an eigenvector of the
interaction Hamiltonian $H_0$ if the coefficient matrix $\m w$ of all
$w_{pq}$ satisfies
\begin{align}
  \bar{\m h}\oc\m w + \m h\c\m w = \f12 \Delta\p{\m w} \m w,
  \label{eq:cond_2}
\end{align}
for some constant $\Delta\p{\m w}$ that may depend on $\m w$.
Remembering that the matrices $\m h$ and $\m w$ are both symmetric
(under transposition) with zeros on the diagonal, the diagonal
components of the condition in \eqref{eq:cond_2} automatically enforce
that the last term ($\sim\v O_{X,Y}$) in \eqref{eq:H_V2_psi} vanishes.

We can recast the condition in \eqref{eq:cond_2} as an eigenvalue
problem by writing the coefficient matrix $\m w$ as a vector
$\v{\m w}$ in $\RR^N\times\RR^N$,
\begin{align}
  \m w = \sum_{p,q} w_{pq} \op{p}{q}
  \to \v{\m w} \equiv \sum_{p,q} w_{pq} \ket{pq},
\end{align}
and defining the matrix
\begin{align}
  \tilde{\m h}
  \equiv \sum_{p,q} \bar h_{pq} \op{pq}
  + \sum_{p,q,\ell} h_{pq} \op{p\ell}{q\ell},
  \label{eq:h_super_mat}
\end{align}
in terms of which the condition in \eqref{eq:cond_2} becomes the
eigenvalue equation
\begin{align}
  \tilde{\m h} \c \v{\m w} = \f12 \Delta\p{\m w} \v{\m w}.
  \label{eq:cond_2_eig}
\end{align}
The matrix $\tilde{\m h}$ in \eqref{eq:h_super_mat} can be written in
the block-diagonal form
\begin{align}
  \tilde{\m h} = \sum_\ell \m h_\ell \otimes \op{\ell},
  &&
  \m h_\ell
  \equiv \sum_{p,q}\p{\delta_{pq} \bar h_{p\ell} + h_{pq}} \op{p}{q},
\end{align}
which reduces the eigenvalue equation in \eqref{eq:cond_2_eig} to the
$N$ smaller, independent eigenvalue equations
\begin{align}
  \m h_\ell \c \v{\m w}_\ell = \f12 \Delta\p{\m w} \v w_\ell,
  &&
  \v w_\ell \equiv \sum_p w_{p\ell} \ket{p}.
  \label{eq:cond_2_eig_block}
\end{align}
We can (numerically) diagonalize each of the blocks $\m h_\ell$, and
classify their eigenvectors $\v w_\ell$ by the corresponding
eigenvalues $\Delta\p{\m w}/2$.  ``Total'' eigenvectors $\v{\m w}$ of
$\tilde{\m h}$ are then built from eigenvectors $\v w_\ell$ within
each block $\m h_\ell$ that share the same eigenvalue
$\Delta\p{\m w}/2$.  The task of constructing eigenvectors
$\V_2\p{\m w_\Delta}\ket\psi$ of $H_0$ and determining their energy
thus reduces to finding the eigenvectors and eigenvalues of $N$
matrices $\m h_\ell$ of size $N\times N$.

@@

We wish to simplify
\begin{align}
  H_0 \V_1 \ket\psi
  = \sum_{\substack{k\\p<q}} h_{pq} v_k \Pi_{pq} X_k \ket\psi,
  \label{eq:H_V1_psi_start_apndx}
\end{align}
which has terms with $k\in\set{p,q}$, and terms with
$k\notin\set{p,q}$.  In the case of $k\notin\set{p,q}$, the
permutation operator $\Pi_{pq}$ commutes with $X_k$ and annihilates on
$\ket\psi$, leaving us the sum
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}} h_{pq}
  = \f12 \sum_{\substack{p,q\\k\notin\set{p,q}}} h_{pq}
  = \f12 \sum_{p,q} h_{pq}
  - \f12 \sum_{\substack{p,q\\k\in\set{p,q}}} h_{pq}
  = E_0 - h_k,
\end{align}
where we have used the facts that $h_{pq}=h_{qp}$ and $h_{pp}=0$.  The
terms in \eqref{eq:H_V1_psi_start_apndx} with $k\notin\set{p,q}$ are
then
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}}
  h_{pq} v_k \Pi_{pq} X_k \ket\psi
  = \sum_k \p{E_0 - h_k} v_k X_k \ket\psi
  = E_0 \V_1 \ket\psi - \sum_k h_k v_k X_k \ket\psi
\end{align}
while the terms with $k\in\set{p,q}$ are
\begin{align}
  \sum_{\substack{p<q\\k\in\set{p,q}}}
  h_{pq} v_k \Pi_{pq} X_k \ket\psi
  = \sum_{k<q} h_{kq} v_k X_q \ket\psi
  + \sum_{p<k} h_{pk} v_k X_p \ket\psi
  = \sum_{k,q} h_{kq} v_k X_q \ket\psi
\end{align}
which implies
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \p{\sum_{p,q} h_{pq} v_p X_q - \sum_q h_q v_q X_q} \ket\psi.
\end{align}

@@

We wish to simplify
\begin{align}
  H_0 \V_2 \ket\psi
  = \f12 \sum_{\substack{k,\ell\\p<q}} w_{k\ell} h_{pq}
  \Pi_{pq} X_k Y_\ell \ket\psi,
  \label{eq:H_V2_psi_start}
\end{align}
which has one term with $\set{p,q}=\set{k,\ell}$, terms with
$p,q\notin\set{k,\ell}$, and terms with $p,q\in\set{k,\ell}$.  The
permutation operator $\Pi_{pq}$ acts trivially on
$X_k Y_\ell \ket\psi$ when $\set{p,q}=\set{k,\ell}$ or
$p,q\notin\set{k,\ell}$, leaving a sum over $p,q$ of the form
\begin{align}
  h_{k\ell} + \sum_{\substack{p<q\\p,q\notin\set{k,\ell}}} h_{pq}
  = \sum_{p<q} h_{pq}
  - \sum_{\substack{p\in\set{k,\ell}\\q\notin\set{k,\ell}}} h_{pq}
  = E_0 + 2 h_{k\ell} - h_k - h_\ell
  = E_0 + 2 \bar h_{k\ell},
\end{align}
where
\begin{align}
  \bar h_{k\ell} \equiv h_{k\ell} - \f12\p{h_k + h_\ell}.
\end{align}
The corresponding terms in \eqref{eq:H_V2_psi_start} with
$p,q\notin\set{k,\ell}$ and $\set{p,q}=\set{k,\ell}$ are then
\begin{align}
  E_0 \V_2 \ket\psi
  + \sum_{k,\ell} \bar h_{k\ell} w_{k\ell} X_k Y_\ell \ket\psi.
\end{align}
The terms in \eqref{eq:H_V2_psi_start} with only one of
$p,q\in\set{k,\ell}$ simplify to
\begin{align}
  \f12 \sum_{\substack{k,\ell\\q\notin\set{k,\ell}}} w_{k\ell}
  \p{h_{kq} X_q Y_\ell + h_{\ell q} X_k Y_q} \ket\psi
  = \sum_{k,\ell,q} w_{k\ell} h_{kq} X_q Y_\ell \ket\psi
  - \sum_{k,\ell} w_{k\ell} h_{k\ell} X_k \odot Y_k \ket\psi
\end{align}
where $A\odot B \equiv \p{AB+BA}/2$ is a symmetrized product.  In
total,
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \sum_{k,\ell} \bar h_{k\ell} w_{k\ell} X_k Y_\ell \ket\psi
  + \sum_{k,\ell,q} h_{kq} w_{q\ell} X_k Y_\ell \ket\psi
  - \sum_{k,\ell} w_{k\ell} h_{k\ell} X_k \odot Y_k \ket\psi.
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Single-body product}
\label{sec:PXXP}
